Free Theorems for Nested Types

### 2020-06-30 free-theorems-nesteds/paper5.tex
small text changes to supp.tex
today's changes to paper5

{\small
\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{Cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]}
A natural approach to the problem uses structural induction on the
input data structure in question. This requires knowing not just the
definition of the ADT of which the input data structure is an
instance, but also the program text for the functions involved in the
properties to be proved. For example, to prove by induction that
mapping a polymorphic function over a list and then reversing the
resulting list is the same as reversing the original list and then
mapping the function over the result, we unwind the (recursive)
definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
over lists to according to the inductive structure of the input
list. Such data-driven induction proofs over ADTs are routine enough
to be included in, say, an undergraduate functional programming
course.

@@

Unfortunately, it is nowhere near as clear that analogous intuitive or
formal inductive arguments can be made for the
$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
by induction on the input bush must recursively induct over the bushes
that are internal to the top-level bush in the recursive position.
This is sufficiently delicate that no induction rule for bushes or
other truly nested types was known until very recently, when {\em deep
  induction}~\cite{jp20} was developed as a way to induct over {\em
  all} of the structured data present in an input. Deep induction thus
not only gave the first principled and practically useful structural
induction rules for bushes and other truly nested types, and has also
opened the way for incorporating automatic generation of such rules
for (truly) nested data types --- and, eventually, even GADTs --- into
modern proof assistants.
\begin{figure*}
\hspace*{-0.5in}
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
  \rightarrow PTree \;A}\\  
\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
\hspace*{1.3in} \mathtt{pleaf}\\
\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
  swap\; p) )}\\  
\\
\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
  \rightarrow}\\
\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F B) \rightarrow}\\
\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F (B \times B)
  \rightarrow F B) \rightarrow}\\
\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
\\
\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
  \rightarrow PLeaves\; A \rightarrow PLeaves\; B}\\ 
\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
\mathtt{mapPTree\; f \;(pnode\; p) = pnode\; (mapPTree \;(\lambda p
  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
\\
\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
  \rightarrow (A \times A)}\\ 
\mathtt{swap \;(x,y) = (y,x)}
\end{array}\]
\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
\end{minipage}}
\quad\quad\quad\quad\quad\quad\quad\quad
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
  \rightarrow Bush\; A}\\ 
\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
\\
\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
  Set\} \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F B) \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
\\
\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to (Bush\, A) \to
  (Bush\, B)}\\ 
\mathtt{mapBush\; \_ \; bnil = bnil}\\
\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
  (mapBush\, f) \; bbx)}\\ 
\\
\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
  \;(Bush\; B) \rightarrow Bush\; B}\\
\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
\end{array}\]
\caption{$\mathtt{reverseBush}$ and auxiliary
  functions in Agda}\label{fig:funs2} 
\end{minipage}}
\end{figure*}

@@

\subsection{Types}
For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
  constructor variables of arity $k$} and $\fvars^k$ of
\emph{functorial variables of arity $k$}, all mutually disjoint.
%disjoint for distinct $k$ and disjoint from each other.
The sets of all type constructor variables and functorial variables
are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
  0} \fvars^k$, respectively, and a \emph{type variable} is any
element of $\tvars \cup \fvars$.  We use lower case Greek letters for
type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
unimportant, or clear from context. We reserve letters from the
beginning of the alphabet to denote type variables of arity $0$, i.e.,
elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
or a set of functorial variables when the cardinality $n$ of the set
is unimportant or clear from context. If $\Pos\,$ is a set of type
variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
omit the vector notation for a singleton set, thus writing $\phi$,
instead of $\overline{\phi}$, for $\{\phi\}$.
\begin{dfn}
Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
  expressions} over $\Pos$ and $V$ are given by
\begin{align*}
  \mcF^\Pos(V) \; ::= \; %\T(V)
  &\mid
\zerot \mid \onet 
\mid \Nat^{\ol{\alpha}} \, \mcF^{\overline{\alpha}}(V) \;
\mcF^{P,\ol{\alpha}}(V) %\mid V \ol{\T(V)}
\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
\mid V\, \ol{\mcF^\Pos(V)}  
\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
\phi^{~k}. \lambda \alpha_1...\alpha_k.  
\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
  \ol{\mcF^{\Pos}(V)}
\end{align*}
\end{dfn}
\noindent
A \emph{type} over $\Pos$ and $V$ is any element of % $\T(V) \cup
$\F^\Pos(V)$.

@@

A {\em type constructor context} is a finite set $\Gamma$ of type
constructor variables, and a {\em functorial context} is a finite set
$\Phi$ of functorial variables. In Definition~\ref{def:wftypes}, a
judgment of the form $\Gamma;\Phi \vdash F$ indicates that the
type $F$ is intended to be functorial in the variables in $\Phi$
but not necessarily in those in $\Gamma$.
\begin{dfn}\label{def:wftypes}
The formation rules for the set $\F \subseteq \bigcup_{V \subseteq
  \tvars, \Pos\, \subseteq \fvars}\F^\Pos(V)$ of\, {\em well-formed
  type expressions} are

@@

\[\begin{array}{cc}
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
%\noindent
%A type $\tau$ is {\em well-formed} if it is
%either a well-formed type constructor expression or a
%well-formed functorial expression.
\end{dfn}
If $F$ is a closed type we may write $\vdash F$, rather than
$\emptyset;\emptyset \vdash F$, for the judgment that it is
well-formed.  Definition~\ref{def:wftypes} ensures that the expected
weakening rules for well-formed types hold, although weakening does
not change the contexts in which types can be formed. If
$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$,
then our rules allow formation of the type $\Gamma;\emptyset \vdash
\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash
\Nat^{\ol \alpha}F\,G$ represents a natural transformation in
$\ol \alpha$ from $F$ to $G$, the type $\Gamma;\emptyset
\vdash \Nat^\emptyset F \,G$ represents the arrow type
$\Gamma \vdash F \to G$ in our calculus. Our calculus
similarly represents the $\forall$-type $\Gamma; \emptyset
\vdash \forall \ol\alpha . F$ as $\Gamma; \emptyset \vdash
\Nat^{\ol\alpha} \,\onet \,F$.
%However, if $\ol\alpha$ is non-empty then $\tau$ cannot be of the
%form $\Nat^{\ol\beta} H \, K$ since $\Gamma; \ol\alpha \vdash
%\Nat^{\ol\beta} H \, K$ is not a valid type judgment in our calculus
%(except by weakening).

@@

In addition to textual replacement, we also have a proper substitution
operation on types. If $F$ is a type over $P$ and $V$, if $\Pos$
and $V$ contain only type variables of arity $0$, and if $k=0$ for
every occurrence of $\phi^k$ bound by $\mu$ in $F$, then we say
that $F$ is {\em first-order}; otherwise we say that $F$ is {\em
  second-order}.  Substitution for first-order types is the usual
capture-avoiding textual substitution. We write $F[\alpha :=
  \sigma]$ for the result of substituting $\sigma$ for $\alpha$ in
$F$, and $F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or
$F[\ol{\alpha := F}]$ when convenient, for $F[\alpha_1 :=
  F_1][\alpha_2 := F_2,...,\alpha_k := F_k]$. Substitution
for second-order types is defined below, where we adopt a similar
notational convention for vectors of types. We note that it is not
correct to substitute along non-functorial variables.

\begin{dfn}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$
%with $k \geq 1$,
and if\, $\Gamma;\Phi, \ol{\alpha} \vdash F$ with
$|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
  F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash H$,
%with $k \geq 1$,
and if\, $\Gamma; \Psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
and $\Phi \cap \Psi = \emptyset$, then $\Gamma,\Psi;\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F]$. Here, the operation $(\cdot)[\phi :=_{\ol \alpha}
  F]$ of {\em second-order type substitution along $\ol\alpha$} is
defined by: 
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
  :=_{\ol{\alpha}} F])\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}.\, G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{dfn}
\noindent
We note that $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with
first-order substitution. We also omit $\ol\alpha$ when convenient.

@@

\[\vdash \fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
  \alpha} \, (\tin_{\beta + \phi(\beta \times \beta)} \circ s) :

@@

$\mathit{swap}$ and $s$ are the terms
\[ \vdash L_{\alpha} p.\, (\pi_2 p, \pi_1 p) :
\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\]

@@

@@

  b}{t' \mapsto  
      \alpha}\, \mathit{swap}\, t'} )

@@

\alpha))\; (\alpha + \mathit{PTree}\,(\alpha \times \alpha))

@@

\[\vdash \fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)}\, balg

@@

\vdash \fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
  \times \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha]}\,
(\map_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta \times

@@

  \mathit{Bush}\,\alpha}\, \tin_{\onet + \beta \times \phi

@@

the type $\emptyset; \gamma \vdash \Nat^\emptyset (\onet + \onet)\,

@@

Let $f: \rho \to \rho'$ for set environments $\rho$ and $\rho'$ (so
that $\rho|_\tvars = \rho'|_\tvars$). The action $\setsem{\Gamma;\Phi
  \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$ on the
morphism $f$ is given as follows:

@@

      \F}\rho}}$.  The latter equality holds because $\rho\phi$ and

@@

  (\map_G^{\overline{K}, \overline{H}} \,\overline{t})_{\ol\gamma} \circ

@@

  (\map_F^{\overline{K}, \overline{H}} \,
  \overline{t})_{\ol\gamma}  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=

@@

L_{\overline{\gamma}} z. (\map_{(\mu \phi. \lambda

@@

    \overline{G}}\, \overline{t})_{\ol\gamma}

@@

    \overline{G}}\, \overline{t})_{\ol\gamma} z)

>>>>

{\small
\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]}
A natural approach to the problem uses structural induction on the
input data structure in question. This requires knowing not just the
definition of the ADT of which the input data structure is an
instance, but also the program text for the functions involved in the
properties to be proved. For example, to prove by induction that
mapping a polymorphic function over a list and then reversing the
resulting list is the same as reversing the original list and then
mapping the function over the result, we unwind the (recursive)
definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
over lists to according to the inductive structure of the input
list. Such data-driven induction proofs over ADTs are routine enough
to be included in, say, an undergraduate functional programming
course.

@@

Unfortunately, it is nowhere near as clear that analogous intuitive or
formal inductive arguments can be made for the
$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
by induction on the input bush must recursively induct over the bushes
that are internal to the top-level bush in the recursive position.
This is sufficiently delicate that no induction rule for bushes or
other truly nested types was known until very recently, when {\em deep
  induction}~\cite{jp20} was developed as a way to induct over {\em
  all} of the structured data present in an input. Deep induction thus
not only gave the first principled and practically useful structural
induction rules for bushes and other truly nested types, and has also
opened the way for incorporating automatic generation of such rules
for (truly) nested data types --- and, eventually, even GADTs --- into
modern proof assistants.
\begin{figure*}
\hspace*{-0.7in}
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
  \rightarrow PTree \;A}\\  
\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
\hspace*{1.3in} \mathtt{pleaf}\\
\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
  swap\; p) )}\\  
\\
\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
  \rightarrow}\\
\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F B) \rightarrow}\\
\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F (B \times B)
  \rightarrow F B) \rightarrow}\\
\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
\\
\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
  \rightarrow PTree\; A \rightarrow PTree\; B}\\ 
\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
\mathtt{mapPTree\; f \;(pnode\; p) =}\\
\hspace*{0.7in}  \mathtt{pnode\; (mapPTree \;(\lambda p
  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
\\
\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
  \rightarrow (A \times A)}\\ 
\mathtt{swap \;(x,y) = (y,x)}
\end{array}\]
\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
\end{minipage}}
\quad\quad\quad\quad\quad%\quad\quad\quad
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
  \rightarrow Bush\; A}\\ 
\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
\\
\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
  Set\} \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F B) \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
\\
\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to Bush\, A \to
  Bush\, B}\\ 
\mathtt{mapBush\; \_ \; bnil = bnil}\\
\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
  (mapBush\, f) \; bbx)}\\ 
\\
\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
  \;(Bush\; B) \rightarrow Bush\; B}\\
\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
\end{array}\]
\caption{$\mathtt{reverseBush}$ and auxiliary
  functions in Agda}\label{fig:funs2} 
\end{minipage}}
\end{figure*}

@@

\subsection{Types}
For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
  constructor variables of arity $k$} and $\fvars^k$ of
\emph{functorial variables of arity $k$}, all mutually disjoint.
%disjoint for distinct $k$ and disjoint from each other.
The sets of all type constructor variables and functorial variables
are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
  0} \fvars^k$, respectively, and a \emph{type variable} is any
element of $\tvars \cup \fvars$.  We use lower case Greek letters for
type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
unimportant, or clear from context. We reserve letters from the
beginning of the alphabet to denote type variables of arity $0$, i.e.,
elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
or a set of functorial variables when the cardinality $n$ of the set
is unimportant or clear from context. If $\Pos\,$ is a set of type
variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
omit the vector notation for a singleton set, thus writing $\phi$,
instead of $\overline{\phi}$, for $\{\phi\}$.
\begin{dfn}
Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
  expressions} over $\Pos$ and $V$ are given by
\begin{align*}
  \mcF^\Pos(V) \; ::= \; %\T(V)
  & \hspace*{0.15in}
\zerot \mid \onet 
\mid \Nat^{\ol{\alpha}} \, \mcF^{\overline{\alpha}}(V) \;
\mcF^{\ol{\alpha}}(V) %\mid V \ol{\T(V)}
\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
\mid V\, \ol{\mcF^\Pos(V)}  
\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
\phi^{~k}. \lambda \alpha_1...\alpha_k.  
\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
  \ol{\mcF^{\Pos}(V)}
\end{align*}
\end{dfn}
\noindent
A \emph{type} over $\Pos$ and $V$ is any element of % $\T(V) \cup
$\F^\Pos(V)$.

@@

A {\em type constructor context} is a finite set $\Gamma$ of type
constructor variables, and a {\em functorial context} is a finite set
$\Phi$ of functorial variables. In Definition~\ref{def:wftypes}, a
judgment of the form $\Gamma;\Phi \vdash F$ indicates that the
type $F$ is intended to be functorial in the variables in $\Phi$
but not necessarily in those in $\Gamma$.
\begin{dfn}\label{def:wftypes}
The formation rules for the set $\F \subseteq \bigcup_{V \subseteq
  \tvars, \Pos\, \subseteq \fvars}\F^\Pos(V)$ of\, {\em well-formed
  types} are

@@

\[\begin{array}{cc}
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
&
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
%\noindent
%A type $\tau$ is {\em well-formed} if it is
%either a well-formed type constructor expression or a
%well-formed functorial expression.
\end{dfn}
If $F$ is a closed type we may write $\vdash F$, rather than
$\emptyset;\emptyset \vdash F$, for the judgment that it is
well-formed.  Definition~\ref{def:wftypes} ensures that the expected
weakening rules for well-formed types hold, although weakening does
not change the contexts in which types can be formed. If
$\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$, then our
rules allow formation of the type $\Gamma;\emptyset \vdash
\Nat^\emptyset F \,G$. Since $\Gamma; \emptyset \vdash \Nat^{\ol
  \alpha}F\,G$ represents a natural transformation in $\ol \alpha$
from $F$ to $G$, the type $\Gamma;\emptyset \vdash \Nat^\emptyset F
\,G$ represents the arrow type $\Gamma \vdash F \to G$ in our
calculus. The type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet
\,F$ similarly represents the $\forall$-type $\Gamma; \emptyset \vdash
\forall \ol\alpha . F$.  Note, however, that
%not even all
%Hindley-Milner types of System F are representable in our calculus.
not all System F types are 
%only those System F types that can be interpreted as functors are
representable in our calculus.
%
%However, if $\ol\alpha$ is non-empty then $\tau$ cannot be of the
%form $\Nat^{\ol\beta} H \, K$ since $\Gamma; \ol\alpha \vdash
%\Nat^{\ol\beta} H \, K$ is not a valid type judgment in our calculus
%(except by weakening).

@@

In addition to textual replacement, we also have a proper substitution
operation on types. If $F$ is a type over $P$ and $V$, if $\Pos$
and $V$ contain only type variables of arity $0$, and if $k=0$ for
every occurrence of $\phi^k$ bound by $\mu$ in $F$, then we say
that $F$ is {\em first-order}; otherwise we say that $F$ is {\em
  second-order}.  Substitution for first-order types is the usual
capture-avoiding textual substitution. We write $F[\alpha :=
  \sigma]$ for the result of substituting $\sigma$ for $\alpha$ in
$F$, and $F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or
$F[\ol{\alpha := F}]$ when convenient, for $F[\alpha_1 :=
  F_1][\alpha_2 := F_2,...,\alpha_k := F_k]$. Substitution
for second-order types is defined below, where we adopt a similar
notational convention for vectors of types. Note that it is not
correct to substitute along non-functorial variables.

\begin{dfn}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$
%with $k \geq 1$,
and if\, $\Gamma;\Phi, \ol{\alpha} \vdash F$ with
$|\ol\alpha| = k$, then $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
  F]$.  Similarly, if \,$\Gamma, \phi^k; \Phi \vdash H$,
%with $k \geq 1$,
and if\, $\Gamma; \ol\psi,\ol{\alpha} \vdash F$ with $|\ol\alpha| = k$
and $\Phi \cap \ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash
H[\phi :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$. Here, the operation
  $(\cdot)[\phi :=_{\ol \alpha} F]$ of {\em second-order type
    substitution along $\ol\alpha$} is defined by:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
  :=_{\ol{\alpha}} F])\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}.\, G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{dfn}
\noindent
We note that $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with
first-order substitution. We also omit $\ol\alpha$ when convenient.

@@

\[\vdash (\fold_{\beta + \phi(\beta \times \beta)}^{\mathit{PTree}\,
  \alpha})_\emptyset \,
%(\tin_{\beta + \phi(\beta \times \beta)} \circ
s :

@@

$\mathit{pleaf}$,$\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms

@@

\vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
L_{\alpha}\, x.\, \inl\, x) \; : \; \Nat^{\alpha}\, \alpha\,
(\mathit{PTree}\,\alpha)\\ 
\vdash \tin_{\beta + \phi (\beta \times \beta)} \circ (
L_{\alpha}\, x.\, \inr\, x) \; : \; \Nat^{\alpha}\, \mathit{PTree}
(\alpha \times \alpha)\,(\mathit{PTree}\,\alpha)\\ 
\vdash L_{\alpha} p.\, (\pi_2 p, \pi_1 p) :
\Nat^{\alpha} (\alpha \times \alpha)\, (\alpha \times \alpha)\\

@@

 \mathit{pleaf}\, b}{t' \mapsto \mathit{pnode}\, 
    (((\map_{\mathit{PTree}\,\alpha}^{\alpha \times \alpha, \alpha \times
      \alpha})_\emptyset\, \mathit{swap})_\alpha\, t')}

@@

\alpha))\; \mathit{PTree}\,\alpha

@@

\[\vdash (\fold^{\mathit{Bush}\,\alpha}_{\onet + \beta \times \phi (\phi \beta)})_\emptyset\, balg

@@

\vdash (\fold_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta
  \times \phi (\phi\beta))[\phi := \mathit{Bush}\,\alpha]})_\emptyset\,
((\map_{\onet + \beta \times \phi (\phi\beta)}^{(\onet + \beta \times

@@

  \mathit{Bush}\,\alpha})_\emptyset\, \tin_{\onet + \beta \times \phi

@@

the type $\emptyset; \gamma \vdash \Nat^\emptyset\, (\mu \beta. \onet + \beta)\,

@@

Let $f: \rho \to \rho'$ be a morphism between set environments $\rho$
and $\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
$\setsem{\Gamma;\Phi \vdash F}f$ of\, $\setsem{\Gamma;\Phi \vdash F}$
on the morphism $f$ is given as follows:

@@

      F}\rho}}$.  The latter equality holds because $\rho\phi$ and

@@

  ((\map_G^{\overline{K}, \overline{H}})_\emptyset \,\overline{t}) \circ

@@

  ((\map_F^{\overline{K}, \overline{H}})_\emptyset \,
  \overline{t})  : \Nat^{\overline{\gamma}} F[\overline{\alpha :=

@@

((\map_{(\mu \phi. \lambda

@@

    \overline{G}})_\emptyset\, \overline{t}) \circ (L_{\overline{\gamma}} z.

@@

    \overline{G}})_\emptyset\, \overline{t})
